Likelihoods for a general class of ARGs under the SMC

### 2025-05-05 review-responses.tex

\begin{point}{\revref} % line 577
 I don't understand ``, along with additional information that allows us to sequentially recover (...) efficiently.'' Recover what?
\end{point}

\reply{
We were missing ``the local trees here'', thanks.
}

%%%%%%%%%%%%%%
\reviewersection{2}
\begin{quote}
This paper does a couple of interrelated things.
\begin{enumerate}
    \item The authors develop a new backward-in-time formulation of the Sequentially Markov Coalescent (SMC) model.
    \item This backward-in-time formulation of the SMC model provides a way to efficiently calculate likelihood for a sample of genomes from this model, provided that we have information on how these genomes are related to each other. The latter is done by a data structure called Ancestral Recombination Graph (ARG). The authors provide an algorithm and Python code for this likelihood calculation (is this emphasised enough across the paper!?).
    \item The authors also evaluate the likelihood under different levels of information available in the ARG. This enables them to connect challenges when comparing methods and outputs from the rigorous, yet ``scale-limited'', probabilistic ARG inference methods with scaleable, yet less-rigorous, heuristic ARG inference methods.
\end{enumerate}
I welcome this paper and quite like it's succinctness. I am not able to follow all the maths, but the rationale and presentation is clear to me and I only have minor comments.
\end{quote}

We're glad the paper found a warm welcome.
And, thanks very much for your attention to detail in the comments.

@@

\begin{point}{\revref} % 121:
 sampled and ancestral haplotypes (haploid genomes), or ...
\end{point}

\reply{
Done.
}

\begin{point}{Fig \ref{fig:smc-unary}:}
 it would be cool if you shade regions (in blue) also in the local trees? That's effectively uncertain time around the A, B, and C node!
\end{point}

\reply{
    Nice idea -- we've done this, and hope it doesn't add too much to the complexity of the figures.
}

@@

\begin{point}{\revref} % 164:
 had to Google what ``abutting'' means! Consider using bordering/adjoning
\end{point}

\reply{
Done.
}

\begin{point}{\revref} % 169:
 I don't fully grasp the meaning of ``tree-valued process'' - tree-generating process or tree-structured process ...
\end{point}

\reply{
We feel that ``tree-valued process'' is a fairly well established term in
more mathematical literature and would prefer to keep it as-is.
}

@@

\begin{point}{\revref} % 387:
 probabilistic inference
\end{point}

\reply{
Done.
}

\begin{point}{\revref} % 392-394:
 I checked the code and is indeed there! Thanks for sharing! The algorithm has an acronym RUNSMC standing for ``Recombination (time) UNaware SMC''. Should this be emphasised in the manuscript? Hmm, but it's not fully unaware, is it? You do consider topology, so think about a better acronym.
\end{point}

\reply{
    Good point, we shouldn't let a nice acronym like this go unexplained.
    We think the algorithm is at least relatively unaware of the times,
    and so we're okay with the current acronym.
}

### 2025-05-05 paper.tex

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{An ARG represented as a graph (top) and as a
series of local trees (bottom). To ensure the one-to-one correspondence
between both representations a node is recorded for each local tree whenever
a node is encountered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary.
Nodes A, B and C are recombinants, i.e, have two parents in the graph
and are unary in the local trees. The blue shaded regions indicate
the range of possible times associated with these recombinants.
The blue arrows show the
standard left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
    with the remaining portion of the local tree following recombination.
% By removing recombinant nodes, we limit the information on the time to each
% of these events to the windows (blue shaded array) delimited by the age of
% the child node and the minimum age of its parents.
}
\label{fig:smc-unary}
\end{figure}

@@

Assuming a randomly mating, diploid population of constant size $N_e$,
common ancestry occurs between
two \emph{overlapping} lineages at rate $\lambda = 1/(2N_e)$.
Here we deviate from the McVean and Cardin formulation,
and define a strict total order on all lineages in $L(t)$:
for each lineage $X_j(t) = [x_{j}, y_{j})$ we define its
order by the left coordinate $x_j$ and label $j$ (to break ties).
The instantaneous rate of coalescence then equals $\lambda$
multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{jk}$,
where
\begin{equation} \label{def:coal}
I_{jk} = \begin{cases}
    1 & X_j > X_k \text{ and } x_j < y_k \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$.
Note that since $X_j > X_k$ implies that $x_j \ge x_k$,
this condition implies that the left endpoint $x_j$ of $X_j$
falls in the interval defined by $X_k$.
From the definition of the strict total order it follows that if $I_{jk} = 1$
then $I_{kj} = 0$.
The newly formed lineage acquires the union of both intervals,
the original lineages are removed, and $L(t)$ is updated accordingly.
% [JK notes: need to explain here that this deliberately introduced
% asymmetry doesn't affect the actual process. Also need to comment on
% how the process terminates. Basically has the same subtleties as McVean and
% Cardin, and not important for our purposes.]
For later use, also define $C_j(t)$ to be the set of
lineages that can coalesce into lineage $j$, i.e.,
$C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$;
thus, $|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$.  \revpoint{1}{2}
Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
which will simplify likelihood computations later. In particular, although
recombination affects \emph{which} lineages can coalesce into $j$, it does not
change the \emph{number} of such lineages: in other words, a recombination
event changes $C_j(t)$ but not $I_{j}(t)$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{ARG likelihood} \label{par:liks}
We have defined the SMC process in terms of rates of various types of events.
In general, the likelihood for a continuous-time Markov process specified in this way
is $\exp(-\Lambda) \prod_i \lambda_i$, \revpoint{1}{3}
where $\lambda_i$ is the rate of the $i^\text{th}$ realised event,
and $\Lambda$ is the sum of all rates of all possible events
integrated over the entire process
(also called the ``total hazard'').
As discussed in the Introduction, in many cases these events are not what is
estimated by inference methods, which produce, more fundamentally, a collection
of genetic inheritance relationships (edges) between ancestral haplotypes (nodes).
Our goal, therefore, is to decompose the overall likelihood into the
per-node and edge contributions, allowing us to compute likelihoods
for the incomplete ancestries estimated by current methods.
% Feel like we need a single sentence here that ties the first and second paragraph together
Using the out- and in-degree of the nodes we'll first reason about the number
of common ancestor events and recombination events that could have resulted in
the observed ARG. Next the information contained in the edges is leveraged to
compute the total hazard associated with both types of events.
% , as well as graceful degradation in the presence of incomplete information.

@@

between time $s$ and $t$ is given by $F(e, s, t) = \int_{s}^{t} I_{e}(z) dz$. \revpoint{1}{4}

@@

this edge simply contributes $\exp(-F(e,t_c,t_p))$ to the likelihood.

@@

All scripts and data used for this manuscript are available at
\url{https://github.com/gertjanbisschop/smc-bit-paper},
and an implementation of the algorithm is available at
\url{https://github.com/gertjanbisschop/runsmc}.
\revpoint{2}{43}

>>>>

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{An ARG represented as a graph (top) and as a
series of local trees (bottom). To ensure the one-to-one correspondence
between both representations a node is recorded for each local tree whenever
a node is encountered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary.
Nodes A, B and C are recombinants, i.e., have two parents in the graph
and are unary in the local trees. The blue shaded regions indicate
the range of possible times associated with these recombinants.
The blue arrows show the
standard left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
    with the remaining portion of the local tree following recombination.
% By removing recombinant nodes, we limit the information on the time to each
% of these events to the windows (blue shaded array) delimited by the age of
% the child node and the minimum age of its parents.
}
\label{fig:smc-unary}
\end{figure}

@@

Assuming a randomly mating, diploid population of constant size $N_e$,
common ancestry occurs between
two \emph{overlapping} lineages at rate $\lambda = 1/(2N_e)$.
Here we deviate from the McVean and Cardin formulation,
and define a strict total order on all lineages in $L(t)$:
for each lineage $X_j(t) = [x_{j}, y_{j})$ we define its
order by the left coordinate $x_j$ and label $j$ (to break ties).
The instantaneous rate of coalescence then equals $\lambda$
multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{j \neq k} I_{jk}$,
where
\begin{equation} \label{def:coal}
I_{jk} = \begin{cases}
    1 & X_j > X_k \text{ and } x_j < y_k \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$.
Note that since $X_j > X_k$ implies that $x_j \ge x_k$,
this condition implies that the left endpoint $x_j$ of $X_j$
falls in the interval defined by $X_k$.
From the definition of the strict total order it follows that if $I_{jk} = 1$
then $I_{kj} = 0$.
The newly formed lineage acquires the union of both intervals,
the original lineages are removed, and $L(t)$ is updated accordingly.
% [JK notes: need to explain here that this deliberately introduced
% asymmetry doesn't affect the actual process. Also need to comment on
% how the process terminates. Basically has the same subtleties as McVean and
% Cardin, and not important for our purposes.]
For later use, also define $I_j(t)$ to be the total rate at which lineages
are coalescing into lineage $j$,
and $C_j(t)$ to be the set of lineages that can coalesce into lineage $j$.
In other words,
$C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$
and $|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
and the total rate of coalescence at time $t$ is $\lambda \sum_{j} |C_j(t)|$.  \revpoint{1}{2}
Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
which will simplify likelihood computations later. In particular, although
recombination affects \emph{which} lineages can coalesce into $j$, it does not
change the \emph{number} of such lineages: in other words, a recombination
event changes $C_j(t)$ but not $I_{j}(t)$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{ARG likelihood} \label{par:liks}
We have defined the SMC process in terms of rates of various types of events.
In general, the likelihood for a continuous-time Markov process specified in this way
is $\exp(-\Lambda) \prod_i \lambda_i$, \revpoint{1}{3}
where $\lambda_i$ is the rate of the $i^\text{th}$ realised event,
and $\Lambda$ is the sum of all rates of all possible events
integrated over the entire process
(also called the ``total hazard'').
Each $\lambda_i$ would be either equal to $\lambda$ (if the event is a coalescence)
or $r|X_j(t)|$ for some $j$ and $t$ (if it is a recombination),
and since the total rates at time $t$ of recombination events is $rT(t)$
and of coalescent events is $\lambda \sum_j |C_j(t)|$,
the total hazard up to time $T$ is $\Lambda = \int_0^T r T(t) + \lambda \sum_j |C_j(t)| dt$.

However,
as discussed in the Introduction, in many cases these events are not what is
estimated by inference methods, which produce, more fundamentally, a collection
of genetic inheritance relationships (edges) between ancestral haplotypes (nodes).
Our goal, therefore, is to decompose the overall likelihood into the
per-node and edge contributions, allowing us to compute likelihoods
for the incomplete ancestries estimated by current methods.
% Feel like we need a single sentence here that ties the first and second paragraph together
Using the out- and in-degree of the nodes we'll first reason about the number
of common ancestor events and recombination events that could have resulted in
the observed ARG. Next the information contained in the edges is leveraged to
compute the total hazard associated with both types of events.
% , as well as graceful degradation in the presence of incomplete information.

@@

between time $s$ and $t$ is given by $\lambda$ multiplied by
$F(e, s, t) = \int_{s}^{t} I_{e}(z) dz$. \revpoint{1}{4}

@@

this edge simply contributes $\exp(-\lambda F(e,t_c,t_p))$ to the likelihood.

@@

All scripts and data used for this manuscript are available at
\url{https://github.com/gertjanbisschop/smc-bit-paper},
and an implementation of the algorithm,
named ``Recombination (time) UNaware SMC'' (runsmc),
is available at
\url{https://github.com/gertjanbisschop/runsmc}.
\revpoint{2}{43}
